\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-01-21 paper/paper.tex
fixes, so compiling Rnw works

Alvaredo (2010) shows formally, how top incomes shares are related to the gini-coefficient. Following his argument, it is crucial to consider top incomes. Leigh (2007) also analyzed the relationship between the Gini (and further inequality measures) and top income shares in a panel of 13 countries (Switzerland is part of the studie). 
%[He also discusses tax data studies in general-> have a closer look] \\
%[We definitely have to argue, why it isn’t sufficient to report top income shares, because if it is, there is no need to have another study about Switzerland (national level>Martinez, cantonal level> gorgas and Schaltenegger)]

@@

\textbf{Summary of tax data drawbacks}
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial “income” see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the “top-income” literature, albeit it is a crucial issue in “theorized” concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see Foellmi and Martinez 2013:11f).(a) only taxed cases are included (partly a problem, for certain periodes we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registred. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. Feld and Frey (2007) report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequaly measures over time.
\end{itemize}

@@

See for other countries: Siminski et al 2003 (Australia), Brewer et al 2008, UK, Burkhauser et al. 2009, US)
%	•	What about the special cases? Foelmi and Martinez argue, that it is crucial to include the special cases because they include the high net wealth individuals taxed according to their expenditures. Exclusion leads to an underestimation of inequality
%	•	Capital gains? Should not be included 
The discussion about problems with reporting income is fairly exhaustive. What about wealth?

@@

\textbf{deductions} Del et al. (2007:477):” we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates.”
Gorgas and Schaltenegger (2011:5): “”..., information on […] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a “perverse redistribution” effect by redistributing income from the lower middle class to the upper middle class (vgl. Peters 2011 and Interpellation Barbara Gysel (2009).

@@

%\subsubsection{Estimating the bias}
%For most of the observed range (1941 to 2010) we do not have any information how many tax units fall into the category of having income that is not zero but is too little to qualify for federal taxation. However starting 1995, the FTA provides exactly this information for each canton. This enables us to estimate the bias we introduce for each canton and each period between 1995 and 2010. Consequently we can obtain information whether the bias is stable over time (which maes it possible to safely interpret the changes of inequality over time) and whether the bias is different for each canton. \textcolor{red}{Unterschiede zwischen Kantonen wären gut um zu argumentieren, dass andere Länder auch davon betroffen sind, in etwa sowas wie "je höher der Steuerfreibetrag, umso stärker der Bias". Länder de erst sehr spät besteuern (und über nicht Besteuerte dann auch nicht Buch führen) haben einen krassen Bias. Wir könnten dann empfehlungen geben, ab welchem Perzentil man save interpretieren kann oder so.}

@@

@@

%Methodological Main findings 
%(1) Tax data is useful because it’s possible to construct consistent measures over time for income and wealth, which is useful to asses changes over time and conducted studies about impact of structural changes.(2) Concerning accuracy of inequality tax data has advantages and shortcomings. It’s superior to survey data, because the latter is affected by non-responded bias. Tax data includes data about the whole population … [was finden wir wegen den Nullern raus] . or at least it should (tax evasion).(3) The FTA-Tax data has different shortcomings, which cannot be handled like a measure of income and wealth which (a) cannot be corrected for household size and (b) is neither a pre-transfer nor a post-transfer income measure. Therefor it’s comparability with other surveys is harsh as long as the main sources of data on inequality are based on measures, which cannot be replicated with taxable income. This FTA-Tax data problem can be handled with non-aggregated tax-data .

>>>>

Alvaredo (2010) shows formally, how top incomes shares are related to the gini-coefficient. Following his argument, it is crucial to consider top incomes. Leigh (2007) also analyzed the relationship between the Gini (and further inequality measures) and top income shares in a panel of 13 countries (Switzerland is part of the studie). 
%[He also discusses tax data studies in general-> have a closer look] \\
%[We definitely have to argue, why it isnt sufficient to report top income shares, because if it is, there is no need to have another study about Switzerland (national level>Martinez, cantonal level> gorgas and Schaltenegger)]

@@

\textbf{Summary of tax data drawbacks}
\begin{itemize}
\item Misreporting of incomes (high earners have an interest in getting tax beneficial income see above)
\item with aggregated data the adjustment for household members is not possible (no equivalization possible). This is a point, which is hardly mentioned in the top-income literature, albeit it is a crucial issue in theorized concepts of income. To generalize one can say, that the concept of tax units (individuals and couples + no direct information about household members at least not in the aggregated tax tables) is not congruent with the concept of household. This might influence the overall inequality, taking into account the change from traditional household and family structures over the last century. This is an inequality related issue, where relevant studies are missing. What are the assumptions?
\item the tax data income definition summarize labor income, capital income and taxable transfer payments (no means-tested benefits as welfare aid). Distinction between the different income sources is not possible, albeit the mechanisms of changes in the specific distribution can be very different.
\item Some tax units are not included in the tax tables (see Foellmi and Martinez 2013:11f).(a) only taxed cases are included (partly a problem, for certain periodes we know how many cases are excluded because of low or none income)	- (b) cases tax at source (Quellenbesteuerung). Foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. In some border-cantons the share of this group is very high (20\%) 
%Can we ignore this)
(c) staff of international organizations based in Switzerland	- (d) non-fillers show up in the tables as long as they are registred. This persons get an imputed income (older tax tax return and information given by employers. Not registered non-	fillers are not in the records.
- (e) tax evasion. Feld and Frey (2007) report about tax evasion in Switzerland, it should be somewhat above 20 percent on average. With the (strong) assumption that the pattern of tax evasion over time is stable, this is a minor problem for inequaly measures over time.
\end{itemize}

@@

See for other countries: Siminski et al 2003 (Australia), Brewer et al 2008, UK, Burkhauser et al. 2009, US)
%		What about the special cases? Foelmi and Martinez argue, that it is crucial to include the special cases because they include the high net wealth individuals taxed according to their expenditures. Exclusion leads to an underestimation of inequality
%		Capital gains? Should not be included 
The discussion about problems with reporting income is fairly exhaustive. What about wealth?

@@

\textbf{deductions} Del et al. (2007:477): we can check with statistics for 1971-72 (as well as later years) presented both by size of income before deductions and income after deductions that adding back deductions does not introduce any significant error in our estimates.
Gorgas and Schaltenegger (2011:5): ..., information on [] deductions is provided in the tax statistics, thus, we could add the personal deductions to the income data to obtain a consistent series over time. Können wir das auch? Zumindest für gewisse Zeiträume? Das wäre noch gut. \\

Studies on income try to focus on the disposable income, which subtracts certain expenditures from the primary income. Deductions reflect somehow compulsory expenditures and thus taxable income can be seen as a sort of pseudo disposable income. On the other hand deductions can affect the distribution. There are recent studies about the correlation of progressivity and deductions in Switzerland, which examines if deductions have a perverse redistribution effect by redistributing income from the lower middle class to the upper middle class (vgl. Peters 2011 and Interpellation Barbara Gysel (2009).

@@

\subsubsection{Estimating the bias}
For most of the observed range (1941 to 2010) we do not have any information how many tax units fall into the category of having income that is not zero but is too little to qualify for federal taxation (lets call those "zeros" for convenience) . However starting 1995, the FTA provides exactly this information for each canton. This enables us to estimate the bias we introduce for each canton and each period between 1995 and 2010. Consequently we can obtain information whether the bias is stable over time (which makes it possible to safely interpret the changes of inequality over time) and whether the bias is different for each canton. \textcolor{red}{Unterschiede zwischen Kantonen wären gut um zu argumentieren, dass andere Länder auch davon betroffen sind, in etwa sowas wie "je höher der Steuerfreibetrag, umso stärker der Bias". Länder de erst sehr spät besteuern (und über nicht Besteuerte dann auch nicht Buch führen) haben einen krassen Bias. Wir könnten dann empfehlungen geben, ab welchem Perzentil man save interpretieren kann oder so.}

One can take a first look at the descriptives plotting the share of zeros over time seperately for each canton .

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/zero_descriptives1}

\includegraphics[width=\maxwidth]{figure/zero_descriptives2}

\end{knitrout}

We can see multiple things here:

\begin{enumerate}
\item There is a small overall upward trend which we assume to be the Federal Administrations inflation adjustments to the tax threshold.
\item Geneva and Tessin show wild changes but those might be explained by the tax gap ("Bemessungslücke") that people exploited when the cantons changed the tax system. It remains unclear however why we can't see similar patter within other cantons.
\item There is some variance and we see different patterns over time and cantons. When estimating gini coefficients or the like we must therefore assume that ignoring "the zeros" leads to a bias that is not stable over time.
\end{enumerate}

We might try two strategies to moderate the problem:

\begin{enumerate}
\item Add the zeros as a seperate group
\item Fit a model to predict the inequality measure (e.g. the gini coefficient) using the share of zeros as a predictor
\end{enumerate}

By adding the zeros as a seperate group we face several problems. With the exception of Geneva 1995/96 Pareto interpolation of percentiles p20 and above seems viable. Low percentiles however would need to be extrapolated so we might impose unrealistic assuptioms. Even when estimating "safe" measures above p20 (better p50) or a gini coefficient we need to make an assumption about the income structure of that group (a distribution, a mean income or zero-income). To be consistent with the measure of "taxable income" we could assume an income of zero for that group, calculate gini coefficients and compare them with the original (uncorrected/plain) version. To get a rough number we could (cantonwise) check the squared correlation between plain and corrected gini.

The second approach however is more robust as we do not impose additional assumptions but instead exercise a little curve fitting.

The model outputs a teststatisic for each canton that tells us whether the variation of the zero-rate over time leads to a significant deviation from the typical "canton gini-level". As the model has a decent fit we are not in great danger of omitted variable bias. Using a joint F-Test we can now test if all canton#zeros interactions are zero.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## Wald test for kanton:zeros
##  in lm(formula = G_steink ~ kanton + kanton:zeros + factor(steuerperiode) + 
##     factor(steuerperiode):zeros, data = df)
## F =  3.282  on  27  and  262  df: p= 4e-07
\end{verbatim}
\end{kframe}
\end{knitrout}

In our case we can clearly reject the hypothesis that all interactions are zero. This leads to the conclusion  that gini coefficients are biased by the variation of the zero-share which is kind of obvious but at the same time we can use the model to report adjusted gini coefficients. For example one might be interested in how inequality would had developed if the zero-share would have been constant over time. (Note RF: predict all data point using canton, time and the initial OR final zero-share to homogenize the time series)
Furthermore we can quantify how large the bias is and we can do this seperately for tax periods or seperately for cantons.

For all cantons:

We can see the the model fit reduces to explaining 61.5\% of the gini variation versus 92.6\% when the information from the zero-shares was used. Although this so some extent attributable to the additional 26 paramters: this is huge.

The model indicates some cases that deserve more attention: Schwyz (positive coefficient) and Geneva (negative coefficient) and the tax period 2000 as well as the most recent periods.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## [1] -0.006575
## [1] -0.005752
## [1] 0.09164
## [1] 0.08027
## [1] 0.07055
## [1] 0.06316
\end{verbatim}
\end{kframe}
\end{knitrout}

A positive coefficient (e.g. Schwyz) can be read as follows: In periods with many zeros we measure higher gini coefficients. Our conclusion is that people with low income are more successfully hidden. If those people would be taxed and therefore appear in the statistic they would sharply contrast the high income earners.

A negative coefficient (e.g. Geneva) means the more zeros there are the smaller the gini measure was compared to other tax periods within that canton (remember this is a fixed-effects model). This is less obvious than the Schwyz case and our reading is: more zeros mask inequality that arises from the bottom.

\subsubsection{Descriptives - where do we put this?}
nice grapics for taxable income, reines einkommen and income after taxes here...
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/different_ginis1}

\includegraphics[width=\maxwidth]{figure/different_ginis2}

\end{knitrout}

@@

%rf: genau. kantone unterstreichen die probleme mit dem 0er bias und die probleme mit "imputing the gap". mehr muss erstmal nicht sein.

@@

%Methodological Main findings 
%(1) Tax data is useful because its possible to construct consistent measures over time for income and wealth, which is useful to asses changes over time and conducted studies about impact of structural changes.(2) Concerning accuracy of inequality tax data has advantages and shortcomings. Its superior to survey data, because the latter is affected by non-responded bias. Tax data includes data about the whole population  [was finden wir wegen den Nullern raus] . or at least it should (tax evasion).(3) The FTA-Tax data has different shortcomings, which cannot be handled like a measure of income and wealth which (a) cannot be corrected for household size and (b) is neither a pre-transfer nor a post-transfer income measure. Therefor its comparability with other surveys is harsh as long as the main sources of data on inequality are based on measures, which cannot be replicated with taxable income. This FTA-Tax data problem can be handled with non-aggregated tax-data .

### 2014-01-21 paper/subdocuments/data_and_methods.Rnw
fixes, so compiling Rnw works

<<setup echo=FALSE>>=
library(foreign)
df <- read.dta("data/ginis_und_perzentile.dta")
@

<<zero_descriptives>>=
library(ggplot2)
df$zeros <- df$null_norm/df$cpop # i only use normal cases here
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+geom_line(aes(y=G),color="blue")
@

@@

<<correct_gini>>=
fit <- lm(G~kanton+kanton:zeros+factor(steuerperiode)+factor(steuerperiode):zeros,data=df)
summary(fit)
@

@@

<<magnitude_of_bias>>=
reduced_fit <- lm(G~kanton+factor(steuerperiode),data=df)
summary(reduced_fit)
@

@@

gini(x02)-gini(na.omit(x02_20))

>>>>

<<setup>>=
library(foreign)
df <- read.dta("../../data/ginis_und_perzentile.dta")
df$zeros <- df$null_norm/df$cpop # i only use normal cases here
@

<<zero_descriptives>>=
library(ggplot2)
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)
ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+geom_line(aes(y=G_steink),color="blue")
@

@@

<<correct_gini>>=
fit <- lm(G_steink~kanton+kanton:zeros+factor(steuerperiode)+factor(steuerperiode):zeros,data=df)
#summary(fit)
@

@@

<<magnitude_of_bias>>=
reduced_fit <- lm(G_steink~kanton+factor(steuerperiode),data=df)
#summary(reduced_fit)
@

@@

library(reldist)
gini(x02)-gini(na.omit(x02_20))
